---
title: R Tutorial
layout: default
---
> This is a beginner, non-programmer friendly guide to learn and understand the R language from scratch, giving a brief walkthrough of the most important parts of the language in plain english, intended to get you on board fairly quick.

I assume that you have R or RStudio installed and are ready to follow along, typing the codes in to your R console as you learn. So without further adieu lets jump right in.

##SECTION 1

####First Touch: R As A Calculator   

Illustration of The Components of R Window

If not for anything else, the R console can be used as a built-in calculator. Open your R console and type the following in. You need not type in anything after the `#` symbol in your console because the hash `#` is a comment character. R ignores everything that comes after the `#`.
 
```{r, eval=FALSE}
2 + 3  # the space around '+' is optional
```
That prints out the number 5 as the answer in your console. Lets try some more.

```{r, eval=FALSE}
2 * 3  #=> 6
sqrt (36)  #=> 6, square root
log10 (100)  #=> 2, log base 10
10 / 3  #=> 3.3, 10 by 3
10 %/% 3  #=> 3, quotient of 10 by 3
10 %% 3  #=> 1, remainder of 10 by 3
```

####The Assignment Operator

The next thing you need to know about is R’s assignment operator. Unlike most other languages, R uses a ‘<-‘ operator in addition to the usual ‘=’ operator for assigning values. So whenever you see a ‘<-‘ in R code, know that it just works like a ‘=’ but in both directions. Here is an example for you to try out in your R console. Alternatively, you can use the R Editor to type in all 4 lines at once and press (Cmd+R) or (Ctrl+R) keys to run the selection or current line.

```{r, eval=FALSE}
a <- 10  # assign 10 to 'a'
a = 10  # same as above
10 -> a  # assign 10 to 'a'
10 = a  # Wrong!. This will try to assign `a` to 10.
```

####Classes or Data types

In previous code, you may have noticed that there is no dedicated step to define the type of variable. R intuitively decides that in the background and assigns a ‘class’ to the variable.

```{r, eval=FALSE}
class (a)  # numeric
```

Based on the value assigned to variable a, R decided to assign its class as a ‘numeric’. If you choose to change it as a character ’10’ instead of number 10, that can be done as follows:

```{r, eval=FALSE}
a <- as.character (a)
print (a)  # prints out the value of a
class (a)  # character
```

Find out what happens when you try to convert a character to a numeric using as.numeric().
The next question naturally is what are the different types of classes available in R. The answer actually is infinite, since the users are free to define new classes, but here is some of most commonly used ones:

####Variable Types

* character  – Strings
* integer  – Integers
* numeric  – Integers + Fractions
* factor  – Categorical variable where each level is a category
* logical  – Boolean
* complex  – Complex numbers


####Data Types

* vector  – A collection of elements of same class
* matrix  – All columns must uniformly contain only one variable type.
* data.frame  – The columns can contain different classes.
* list  – Can hold objects of different classes and lengths  

***

###Some Miscellaneous but Important Items Before We Proceed . .

####What is a R package and how to install them?

Upon your first install, R comes with a built-in set of packages which can be invoked directly from your R console. However, since R is a open-source language, anyone can contribute to its capabilities by writing packages. Over the years, these contributions have resulted in growing list of over 5k packages. Here is how you can install the packages from within R console:

```{r, eval=FALSE}
install.packages ("car")  # install car package 
```

The above code will prompt you to select the CRAN mirror for which you can select one that is closest to your place. The dot (.) in ‘install.packages’ is a part of its name and not two separate commands.

Now that the package is installed, you need to initialise it before you can call the functions and datasets that come with the installed package.

```{r, eval=FALSE}
library (car)  # initialise the pkg 'car'
require (car)  # another way to initialise
library ()  # see list of all installed packages
library (help=car)  # see info about 'car' pkg
```


####Getting Help

The easiest way to get help in R is using the ‘?’ operator. Just append a ‘?’ before the name of a function you want to get help, R will open find information about the function from the set of installed packages. If you want to search for it outside the installed packages, use ‘??’ before the function name. ?? can also help search for partial and incomplete terms.
```{r, eval=FALSE}
help (merge)  # get help page for 'merge'
?merge  # lookup 'merge' from installed pkgs
??merge  # vague search
example (merge)  # show code examples
```

####What is a working directory and how to set up one?

A working directory is the reference directory from which R has direct access to read in files. You can read in and write files directly to the working directory without using the full file path. The directory names should be separated by forward slash (/) or double back slash (\\) instead of (\) even for a windows PC.

```{r, eval=FALSE}
getwd ()  # gets the working directory
setwd (dirname)  # set the working directory to dirname
```

####How to import and export data?

The most common and convenient way to bring in data to R is through .csv files. There are packages to import data from excel files (.xlsx) and databases, but that will not be covered here.
```{r, eval=FALSE}
myData <- read.table ("c:/myInputData.txt", header = FALSE, sep="|", colClasses=c("integer","character","numeric") # import "|" separated .txt file
```

```{r, eval=FALSE}
myData <- read.csv ("c:/myInputData.csv", header=FALSE)  # import csv file
write.csv (rDataFrame, "c:/output.csv")  # export 
```

R will intuitively find out what data type the columns in a data frame should be assigned. If you want to manually assign it, it can be set with the ‘colClasses’ argument within read.csv(), which is in fact recommended as it improves the efficiency of the import process.









