# ggplot2 Quickref

#### Basics tasks
* [Basic plot setup](ggplot2-cheatsheet.html#Basic plot setup)
* [Scatterplot](ggplot2-cheatsheet.html#Scatterplot)
* [Static - point size, shape, color and boundary thickness](ggplot2-cheatsheet.html#Static - point size, shape, color and boundary thickness)
* [Dynamic - point size, shape, color and boundary thickness](ggplot2-cheatsheet.html#Dynamic - point size, shape, color and boundary thickness)
* [Add title, X and Y axis labels](ggplot2-cheatsheet.html#Add Title, X and Y axis labels)
* [Change color of all text](ggplot2-cheatsheet.html#Change color of all text)
* [Change title, X and Y axis label size](ggplot2-cheatsheet.html#Change title, X and Y axis label size)
* [Change title face, color, line height](ggplot2-cheatsheet.html#Change title face, color, line height)
* [Change point color](ggplot2-cheatsheet.html#Change point color)
* [Adjust X and Y axis limits](ggplot2-cheatsheet.html#Adjust X and Y axis limits)
* [Change X and Y axis labels](ggplot2-cheatsheet.html#Change X and Y axis labels)
* [Rotate axis text](ggplot2-cheatsheet.html#Rotate axis text)
* [Flip X and Y axis](ggplot2-cheatsheet.html#Flip X and Y Axis)
* [Grid lines and panel background](ggplot2-cheatsheet.html#Grid lines and panel background)
* [Plot margin and background](ggplot2-cheatsheet.html#Plot margin and background)
* [Colors](ggplot2-cheatsheet.html#Colors)

#### Legend
* [Hide legend](ggplot2-cheatsheet.html#Hide legend)
* [Change legend title](ggplot2-cheatsheet.html#Change legend title)
* [Change legend and point color](ggplot2-cheatsheet.html#Change legend and point color)
* [Change legend position](ggplot2-cheatsheet.html#Change legend position)
* [Change order of legend items](ggplot2-cheatsheet.html#Change order of legend items)
* [Change legend title, text, box, symbol](ggplot2-cheatsheet.html#Legend title, text, box, symbol)

#### Plot text and annotation
* [Add text in chart](ggplot2-cheatsheet.html#Add text in chart)
* [Annotation](ggplot2-cheatsheet.html#Annotation)


#### Multiple plots
* [Multiple chart panels](ggplot2-cheatsheet.html#Multiple chart panels)
* [Free X and Y axis scales](ggplot2-cheatsheet.html#Free X and Y axis scales)
* [Arrange multiple plots](ggplot2-cheatsheet.html#Arrange multiple plots)

#### Geom layers
* [Add smoothing line](ggplot2-cheatsheet.html#Add smoothing line)
* [Add horizontal / vertical line](ggplot2-cheatsheet.html#Add horizontal / vertical line)
* [Add bar chart](ggplot2-cheatsheet.html#Add bar chart)
* [Distinct color for bars](ggplot2-cheatsheet.html#Distinct color for bars)
* [Change color and width of bars](ggplot2-cheatsheet.html#Change color and width of bars)
* [Change color palette](ggplot2-cheatsheet.html#Change color palette)
* [Line chart](ggplot2-cheatsheet.html#Line chart)
* [Line chart from timeseries](ggplot2-cheatsheet.html#Line chart from timeseries)
* [Ribbons](ggplot2-cheatsheet.html#Ribbons)
* [Area](ggplot2-cheatsheet.html#Area)
* [Boxplot and Violin](ggplot2-cheatsheet.html#Boxplot and Violin)
* [Density](ggplot2-cheatsheet.html#Density)
* [Tiles](ggplot2-cheatsheet.html#Tiles)

Not finding what you were looking for? [Let me know!](https://docs.google.com/forms/d/13GrkCFcNa-TOIllQghsz2SIEbc-YqY9eJX02B19l5Ow/viewform)

## Basic tasks

#### Basic plot setup

```{r, eval=FALSE}
gg <- ggplot(df, aes(x=xcol, y=ycol)) 
```

`df` must be a dataframe that contains all information to make the ggplot. Plot will show up only after adding the geom layers.

#### Scatterplot

```{r, eval=FALSE}
library(ggplot2)
gg <- ggplot(diamonds, aes(x=carat, y=price)) 
gg + geom_point()
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_1.png", dpi = 220)  
```

#### Static - point size, shape, color and boundary thickness

```{r, eval=FALSE}  
gg + geom_point(size=1, shape=1, color="steelblue", stroke=2)  # 'stroke' controls the thickness of point boundary
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_2.png", dpi = 220)  
```

#### Dynamic - point size, shape, color and boundary thickness

Make the aesthetics vary based on a variable in `df`.
```{r, eval=FALSE}    
gg + geom_point(aes(size=carat, shape=cut, color=color, stroke=carat))  # carat, cut and color are variables in `diamonds`       
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_3.png", dpi = 220)  
```

#### Add Title, X and Y axis labels

```{r, eval=FALSE}    
gg1 <- gg + geom_point(aes(color=color))
gg2 <- gg1 + labs(title="Diamonds", x="Carat", y="Price")  # ggtitle("title") also changes the title.
print(gg2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_4.png", dpi = 220)  
```

#### Change color of all text

```{r, eval=FALSE}  
gg2 + theme(text=element_text(color="blue"))  # all text turns blue.
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_5.png", dpi = 220)  
```

#### Change title, X and Y axis label and text size

`plot.title`: Controls plot title.
`axis.title.x`: Controls X axis title
`axis.title.y`: Controls Y axis title
`axis.text.x`: Controls X axis text
`axis.text.y`: Controls y axis text

```{r, eval=FALSE}
gg3 <- gg2 + theme(plot.title=element_text(size=25), axis.title.x=element_text(size=20), axis.title.y=element_text(size=20), axis.text.x=element_text(size=15), axis.text.y=element_text(size=15))
print(gg3)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_6.png", dpi = 220)  
```

#### Change title face, color, line height

```{r, eval=FALSE}
gg3 + labs(title="Plot Title\nSecond Line of Plot Title") + theme(plot.title=element_text(face="bold", color="steelblue", lineheight=1.2))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_7.png", dpi = 220)  
```

#### Change point color
```{r, eval=FALSE}  
gg3 + scale_colour_manual(name='Legend', values=c('D'='grey', 'E'='red', 'F'='blue', 'G'='yellow', 'H'='black', 'I'='green', 'J'='firebrick'))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_8.png", dpi = 220)  
```

#### Adjust X and Y axis limits

Method 1: Zoom in
```{r, eval=FALSE}
gg3 + coord_cartesian(xlim=c(0,3), ylim=c(0, 5000)) + geom_smooth()  # zoom in
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_9.png", dpi = 220)  
```

Method 2: Deletes the points outside limits
```{r, eval=FALSE}
gg3 + xlim(c(0,3)) + ylim(c(0, 5000)) + geom_smooth()  # deletes the points 
#> Warning messages:
#> 1: Removed 14714 rows containing non-finite values (stat_smooth). 
#> 2: Removed 14714 rows containing missing values (geom_point). 
```    
    
Method 3: Deletes the points outside limits
```{r, eval=FALSE}
gg3 + scale_x_continuous(limits=c(0,3)) + scale_y_continuous(limits=c(0, 5000)) + geom_smooth()  # deletes the points outside limits
#> Warning message:
#> Removed 14714 rows containing missing values (geom_point). 
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_11.png", dpi = 220)  
```

Notice the change in smoothing line because of deleted points. This could sometimes be misleading in your analysis.

#### Change X and Y axis labels
```{r, eval=FALSE}
gg3 + scale_x_continuous(labels=c("zero", "one", "two", "three", "four", "five")) + scale_y_continuous(breaks=seq(0, 20000, 4000))  # if Y is continuous, if X is a factor
```

Use `scale_x_discrete` instead, if X variable is a factor.

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_12.png", dpi = 220)  
```

#### Rotate axis text

```{r, eval=FALSE}
gg3 + theme(axis.text.x=element_text(angle=45), axis.text.y=element_text(angle=45))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_13.png", dpi = 220)  
```

#### Flip X and Y Axis
```{r, eval=FALSE}
gg3 + coord_flip()  # flips X and Y axis.
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_14.png", dpi = 220)  
```

#### Grid lines and panel background
```{r, eval=FALSE}
gg3 + theme(panel.background = element_rect(fill = 'springgreen'),
  panel.grid.major = element_line(colour = "firebrick", size=3),
  panel.grid.minor = element_line(colour = "blue", size=1))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_36.png", dpi = 220)  
```

#### Plot margin and background
```{r, eval=FALSE}
gg3 + theme(plot.background=element_rect(fill="yellowgreen"), plot.margin = unit(c(2, 4, 1, 3), "cm")) # top, right, bottom, left
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_37.png", dpi = 220)  
```

#### Colors

The whole list of colors are displayed at your R console in the `color()` function. Here are few of my suggestions for nice looking colors and backgrounds:

* steelblue  (points and lines)
* firebrick (point and lines)
* springgreen (fills)
* violetred (fills)
* tomato (fills)
* skyblue (bg)
* sienna (points, lines)
* slateblue (fills)
* seagreen (points, lines, fills)
* sandybrown (fills)
* salmon (fills)
* saddlebrown (lines)
* royalblue (fills)
* orangered (point, lines, fills)
* olivedrab (points, lines, fills)
* midnightblue (lines)
* mediumvioletred (points, lines, fills)
* maroon (points, lines, fills)
* limegreen (fills)
* lawngreen (fills)
* forestgreen (lines, fills)
* dodgerblue (fills, bg)
* dimgray (grids, secondary bg)
* deeppink (fills)
* darkred (lines, points)

If you are looking for consistent colors, the `RColorBrewer` package has predefined [color palettes](http://127.0.0.1:4000/ggplot2-cheatsheet.html#Change%20color%20palette)

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_sp37.png", dpi = 220)  
```

```{r, eval=FALSE, echo=FALSE}
colors <- c("whitesmoke", "steelblue", "firebrick", "springgreen", "violetred", "tomato", "thistle", "skyblue", 
            "sienna", "slateblue", "seashell", "seagreen", "sandybrown", "salmon", "saddlebrown", "royalblue", "rosybrown", "powderblue", "plum", "palegoldenrod", "orangered", "olivedrab", "navyblue",
            "navajowhite", "midnightblue", "mediumvioletred", "maroon", "limegreen", "lightgoldenrod",
            "lawngreen", "forestgreen", "dodgerblue", "dimgray", "deeppink", "darkred", "darkkhaki", "azure")
df <- data.frame(num=rep(10, 37), cols=factor(colors, ordered=T))
gg_bar <- ggplot(df, aes(x=cols, y=num)) + geom_bar(stat = "identity", width = 0.5, fill=df$cols) + theme_bw() + theme(axis.text.x=element_text(angle=65, size=15), plot.title=element_text(size=30)) + scale_x_discrete(labels=df$cols) + labs(title="Special 37 Colors", x="", y="")
print(gg_bar)
```

## Legend

#### Hide legend
```{r, eval=FALSE}  
gg3 + theme(legend.position="none")  # hides the legend
```

#### Change legend title
```{r, eval=FALSE}  
gg3 + scale_color_discrete(name="")  # Remove legend title (method1)
p1 <- gg3 + theme(legend.title=element_blank())  # Remove legend title (method)
p2 <- gg3 + scale_color_discrete(name="Diamonds")  # Change legend title
library(gridExtra)
grid.arrange(p1, p2, ncol=2)  # arrange
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_15.png", dpi = 220)  
```

#### Change legend and point color
```{r, eval=FALSE}  
gg3 + scale_colour_manual(name='Legend', values=c('D'='grey', 'E'='red', 'F'='blue', 'G'='yellow', 'H'='black', 'I'='green', 'J'='firebrick'))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_16.png", dpi = 220)  
```

#### Change legend position
__Outside plot__
```{r, eval=FALSE}
p1 <- gg3 + theme(legend.position="top")  # top / bottom / left / right
```    

__Inside plot__
```{r, eval=FALSE}
p2 <- gg3 + theme(legend.justification=c(1,0), legend.position=c(1,0))  # legend justification is the anchor point on the legend, considering the bottom left of legend as (0,0)
gridExtra::grid.arrange(p1, p2, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_17.png", dpi = 220)  
```

#### Change order of legend items
```{r, eval=FALSE}
df$newLegendColumn <- factor(df$legendcolumn, levels=c(new_order_of_legend_items), ordered = TRUE) 
```
Create a new factor variable used in the legend, ordered as you need. Then use this variable instead in the plot.

#### Legend title, text, box, symbol
* `legend.title` - Change legend title
* `legend.text` - Change legend text
* `legend.key` - Change legend box
* `guides` - Change legend symbols

```{r, eval=FALSE}  
gg3 + theme(legend.title = element_text(size=20, color = "firebrick"), legend.text = element_text(size=15), legend.key=element_rect(fill='steelblue')) + guides(colour = guide_legend(override.aes = list(size=2, shape=4, stroke=2)))  # legend title color and size, box color, symbol color, size and shape.
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_18.png", dpi = 220)  
```

```{r, echo = FALSE}
embed_png("screenshots/symbols.png", dpi = 220)  
```


## Plot text and annotation

#### Add text in chart
```{r, eval=FALSE}  
#> Not Run: gg + geom_text(aes(xcol, ycol, label=round(labelCol), size=3))  # general format
gg + geom_text(aes(label=color, color=color), size=4) 
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_19.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Annotation
```{r, eval=FALSE}  
#> gg3 + annotate("mytext", x=xpos, y=ypos, label="My text")  # Not run: General Format
library(grid)
my_grob = grobTree(textGrob("My Custom Text", x=0.8, y=0.2, gp=gpar(col="firebrick", fontsize=25, fontface="bold")))
gg3 + annotation_custom(my_grob)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_20.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

## Multiple plots

#### Multiple chart panels
```{r, eval=FALSE}  
p1 <- gg1 + facet_grid(color ~ cut)  # arrange in a grid. More space for plots.
```

#### Free X and Y axis scales

By setting `scales='free'`, the scales of both X and Y axis is freed. Use `scales='free_x'` to free only X-axis and `scales='free_y'` to free only Y-axis.

```{r, eval=FALSE}  
p2 <- gg1 + facet_wrap(color ~ cut, scales="free")  # free the x and y axis scales.
```

#### Arrange multiple plots
```{r, eval=FALSE}    
library(gridExtra)
grid.arrange(p1, p2, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_21.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

## Geom layers

#### Add smoothing line
```{r, eval=FALSE}
gg3 + geom_smooth(aes(color=color))  # method could be - 'lm', 'loess', 'gam'
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_22.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Add horizontal / vertical line
```{r, eval=FALSE}
p1 <- gg3 + geom_hline(yintercept=5000, size=2, linetype="dotted", color="blue") # linetypes: solid, dashed, dotted, dotdash, longdash and twodash
p2 <- gg3 + geom_vline(xintercept=4, size=2, color="firebrick")
p3 <- gg3 + geom_segment(aes(x=4, y=5000, xend=4, yend=10000, size=2, lineend="round"))
p4 <- gg3 + geom_segment(aes(x=carat, y=price, xend=carat, yend=price-500, color=color), size=2) + coord_cartesian(xlim=c(3, 5))  # x, y: start points. xend, yend: end points
gridExtra::grid.arrange(p1,p2,p3,p4, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_23.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Add bar chart
```{r, eval=FALSE}
# Frequency bar chart: Specify only X axis.
gg <- ggplot(mtcars, aes(x=cyl))
gg + geom_bar()  # frequency table
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_24.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

```{r, eval=FALSE}
gg <- ggplot(mtcars, aes(x=cyl))
p1 <- gg + geom_bar(position="dodge", aes(fill=factor(vs)))  # side-by-side
p2 <- gg + geom_bar(aes(fill=factor(vs)))  # stacked
gridExtra::grid.arrange(p1, p2, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_25.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

```{r, eval=FALSE}
# Absolute bar chart: Specify both X adn Y axis. Set stat="identity"
df <- aggregate(mtcars$mpg, by=list(mtcars$cyl), FUN=mean)  # mean of mpg for every 'cyl'
names(df) <- c("cyl", "mpg")
head(df)
#>   cyl    mpg
#> 1   4  26.66
#> 2   6  19.74
#> 3   8  15.10

gg_bar <- ggplot(df, aes(x=cyl, y=mpg)) + geom_bar(stat = "identity")  # Y axis is explicit. 'stat=identity'
print(gg_bar)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_26.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Distinct color for bars
```{r, eval=FALSE}    
gg_bar <- ggplot(df, aes(x=cyl, y=mpg)) + geom_bar(stat = "identity", aes(fill=cyl))
print(gg_bar)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_27.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Change color and width of bars
```{r, eval=FALSE}    
df$cyl <- as.factor(df$cyl)
gg_bar <- ggplot(df, aes(x=cyl, y=mpg)) + geom_bar(stat = "identity", aes(fill=cyl), width = 0.25)
gg_bar + scale_fill_manual(values=c("4"="steelblue", "6"="firebrick", "8"="darkgreen"))
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_28.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Change color palette
```{r, eval=FALSE}    
library(RColorBrewer)
display.brewer.all(n=20, exact.n=FALSE)  # display available color palettes
ggplot(mtcars, aes(x=cyl, y=carb, fill=factor(cyl))) + geom_bar(stat="identity") + scale_fill_brewer(palette="Reds")  # "Reds" is palette name
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_29.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Line chart
```{r, eval=FALSE}  
# Method 1:
gg <- ggplot(economics, aes(x=date))  # setup
gg + geom_line(aes(y=psavert), size=2, color="firebrick") + geom_line(aes(y=uempmed), size=1, color="steelblue", linetype="twodash")  # No legend
# available linetypes: solid, dashed, dotted, dotdash, longdash and twodash
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_30.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

```{r, eval=FALSE}  
# Method 2:
library(reshape2)
df_melt <- melt(economics[, c("date", "psavert", "uempmed")], id="date")  # melt by date. 
gg <- ggplot(df_melt, aes(x=date))  # setup
gg + geom_line(aes(y=value, color=variable), size=1) + scale_color_discrete(name="Legend")  # gets legend.
```  

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_31.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Line chart from timeseries
```{r, eval=FALSE}
# One step method.
library(ggfortify)
autoplot(AirPassengers, size=2) + labs(title="AirPassengers")
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_32.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Ribbons
Filled time series can be plotted using `geom_ribbon()`. It takes two compulsory arguments `ymin` and `ymax`.
```{r, eval=FALSE} 
# Prepare the dataframe
st_year <- start(AirPassengers)[1]
st_month <- "01"
st_date <- as.Date(paste(st_year, st_month, "01", sep="-"))
dates <- seq.Date(st_date, length=length(AirPassengers), by="month")
df <- data.frame(dates, AirPassengers, AirPassengers/2)
head(df)
#>        dates AirPassengers AirPassengers.2
#> 1 1949-01-01           112            56.0
#> 2 1949-02-01           118            59.0
#> 3 1949-03-01           132            66.0
#> 4 1949-04-01           129            64.5
#> 5 1949-05-01           121            60.5
#> 6 1949-06-01           135            67.5
# Plot ribbon with ymin=0
gg <- ggplot(df, aes(x=dates)) + labs(title="AirPassengers") + theme(plot.title=element_text(size=30), axis.title.x=element_text(size=20), axis.text.x=element_text(size=15))
gg + geom_ribbon(aes(ymin=0, ymax=AirPassengers)) + geom_ribbon(aes(ymin=0, ymax=AirPassengers.2), fill="green")
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_33.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

```{r, eval=FALSE} 
gg + geom_ribbon(aes(ymin=AirPassengers-20, ymax=AirPassengers+20)) + geom_ribbon(aes(ymin=AirPassengers.2-20, ymax=AirPassengers.2+20), fill="green")
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_34.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```


#### Area
`geom_area` is similar to `geom_ribbon`, except that the `ymin` is set to 0. If you want to make overlapping area plot, use the `alpha` aesthetic to make the top layer translucent.
```{r, eval=FALSE} 
# Method1: Non-Overlapping Area
df <- reshape2::melt(economics[, c("date", "psavert", "uempmed")], id="date")
head(df, 3)
#>         date variable value
#> 1 1967-07-01  psavert  12.5
#> 2 1967-08-01  psavert  12.5
#> 3 1967-09-01  psavert  11.7
p1 <- ggplot(df, aes(x=date)) + geom_area(aes(y=value, fill=variable)) + labs(title="Non-Overlapping - psavert and uempmed")

# Method2: Overlapping Area
p2 <- ggplot(economics, aes(x=date)) + geom_area(aes(y=psavert), fill="yellowgreen", color="yellowgreen") + geom_area(aes(y=uempmed), fill="dodgerblue", alpha=0.7, linetype="dotted") + labs(title="Overlapping - psavert and uempmed")
gridExtra::grid.arrange(p1, p2, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_35.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Boxplot and Violin
The oulier points are controlled by the following aesthetics:
* outlier.shape
* outlier.stroke
* outlier.size
* outlier.colour

If the `notch` is turned on (by setting it TRUE), the below boxplot is produced. Else, you would get the standard rectangular boxplots.

```{r, eval=FALSE} 
p1 <- ggplot(mtcars, aes(factor(cyl), mpg)) + geom_boxplot(aes(fill = factor(cyl)), width=0.5, outlier.colour = "dodgerblue", outlier.size = 4, outlier.shape = 16, outlier.stroke = 2, notch=T) + labs(title="Box plot")  # boxplot
p2 <- ggplot(mtcars, aes(factor(cyl), mpg)) + geom_violin(aes(fill = factor(cyl)), width=0.5, trim=F) + labs(title="Violin plot (untrimmed)")  # violin plot
gridExtra::grid.arrange(p1, p2, ncol=2)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_38.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Density

```{r, eval=FALSE} 
ggplot(mtcars, aes(mpg)) + geom_density(aes(fill = factor(cyl)), size=2) + labs(title="Density plot")  # Density plot
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_39.png", dpi = 220)  # legend title color and size, box color, symbol color, size and shape.
```

#### Tiles
```{r, eval=FALSE}
corr <- round(cor(mtcars), 2)
df <- reshape2::melt(corr)
gg <- ggplot(df, aes(x=Var1, y=Var2, fill=value, label=value)) + geom_tile() + theme_bw() + geom_text(aes(label=value, size=value), color="white") + labs(title="mtcars - Correlation plot") + theme(text=element_text(size=20), legend.position="none")

library(RColorBrewer)
p2 <- gg + scale_fill_distiller(palette="Reds")
p3 <- gg + scale_fill_gradient2()
gridExtra::grid.arrange(gg, p2, p3, ncol=3)
```

```{r, echo = FALSE}
embed_png("screenshots/gg_cs_40.png", dpi = 220)
```

